---
title: "Assignment 4"
subtitle: "STAE04/DABN19: Data Visualization"
author: "Stephany Rojas Gerena"
date: "`r Sys.Date()`"
papersize: a4
geometry: margin=4cm
colorlinks: true
output:
  pdf_document:
    number_sections: true
---

```{r setup, include = FALSE}
# Setup options for R Markdown
knitr::opts_chunk$set(
  echo       = FALSE,    # Do not print code
  warning    = FALSE,    # Suppress warnings
  message    = FALSE,    # Suppress messages
  fig.align  = "center", # Center figures
  fig.width  = 2.7,      # Good standard figure width for single-panel figures
  fig.height = 2.4       # Good standard figure height for single-panel figures
)

library(tidyverse)
library(MASS)

# Set a theme for ggplot2
theme_set(theme_grey(base_size = 10))

# Set options
options(
  digits = 3, # limit the number of significant digits
  width  = 63 # limit the width of code output
)
```

# Task 1


We will use a data set containing a study of 2287 eighth-grade pupils (aged about 11) in 132 classes in 131 schools in the Netherlands in 1999. In table 1 we can see a detailed description of the original variables of the data set.

Table 1. Variables Description

| Variable |Type  |Description |
|:----------|:------------|:------------|
| lang | integer |language test score |
| IQ | number |verbal IQ |
| class | factor |class ID |
| GS  | integer | class size: number of eighth-grade pupils recorded in the class |
| SES | integer |social-economic status of pupil's family |
| COMB | factor |were the pupils taught in a multi-grade class (0/1)? |


We are interested in analyzing associations between class size, the type of class (single or multi-grade), and the score on the language test. With that in mind, we will use a scatter plot with class size in the x axis and language test score in the y axis. To avoid overlapping we use jitter, and we also add opacity setting alpha=0.3. Since we want to compare the type of class, we will use facet wrap to have one plot per class. In order to see the trend we will use geom_smooth.


In figure 1 we can see the scatter plot of language test scores vs the size of the class for each type of class. Comparing the type of class the first thing we can see is that in multi-grade the dispersion is higher than in single-grade in general across all size classes. Also, we can see that we have more observations for single-grade because we have a significantly higher amount of dots that cause more overlap compared to the multi-grade graph. By looking at the trend line we can see that single-grade class has higher language test scores in general compared to multi-grade, and also less dispersion. So it is better for students to be in a single-grade class. 

For multi-grade, the relationship between the size of the class and the language test score is positive, so as we increase the size class, the language test score also increase. However, the dispersion is bigger in small size classes, and in big size class compared to sizes between 20-30. So we can say that for multigrade the ideal size of the class is between 20 and 30 students in order to have less dispersion but not entirely sacrifying the score results. 

For single grade the relationship seems very flat, even a little inverse. Also, the dispersion does not seem to be significantly higher in any class size. So in general we could say that the size of the class does not have a significant relationship with the language test score. 


```{r,fig.height=5,fig.width=7,fig.cap="Language test scores across size class and grade of students in Neteherlands"}

library(tidyverse)
data("nlschools", package = "MASS")

nlschools2 <-
  as_tibble(nlschools) %>%
  mutate(COMB = fct_recode(COMB, "single-grade" = "0", "multi-grade" = "1")) %>%
  rename(type_class="COMB",class_size="GS")


ggplot(nlschools2,aes(class_size,lang),)+geom_point(position=position_jitter(),alpha=0.3)+facet_wrap("type_class")+geom_smooth(formula=y~x,method = "lm",color="red")+labs(x="Size class",y="Language test score")+ theme(axis.title=element_text(size=11,face="bold"))


```

# Task 2

After having analyzing the data in the previous task, we see that there is something we are ignoring, and is the fact that the observations belong to different classes. For taking this into account, we will group the data by class, class size and type of class. Also, we will use the mean of each group to plot a scatter plot. We will also use geom_smooth to add a linear trend line that makes the interpretation easier. 


In figure 2 we can see the scatter plot of the mean per group of language test score vs the size of the class per type of class. Here each dot will represent a group of students. We can see that compared to the previous task, the relationships are more pronounced here. We can see that the dispersion is more pronounced for both type of classes compared to the previous task. The dispersion is higher in multi-grade than in single-grade. Overall, we can see that single-grade class performs better than multi-grade for all class sizes. 


In we have a closer look to each type class we see that the relationship between the mean of language score and the size of class is negative for single-grade and positive for multi-grade. It means that having more students in the class increases the mean of the score in multi-grade, but decreases the score in single-grade. Despite this fact, single-grade is always preferable since it has lower dispersion and higher scores compared to multi-grade. 


The main disadvantages of this type of grouping analysis are:

1. We are loosing information about the outlines. In fact, outlines can be adding noise to our groups.
2. There could be a bias by applying regression techniques to grouped data (in our case the red line of the linear regression). This is because the summary measures does not always reflect the behavior of the individuals who make up the group
3. Sometimes relationships change when comparing group data vs. the individual data. So if we are going to group it is always important to look at the non grouped data as well.


```{r,fig.height=5,fig.width=7,fig.cap="Mean language test scores across size class and type of class per group of students in Neteherlands"}

nlschools3 <- nlschools2 %>% group_by(class,class_size,type_class)


ggplot(nlschools3,aes(class_size,lang))+facet_wrap("type_class")+stat_summary_bin(fun = mean,geom = "point")+geom_smooth(formula=y~x,method="lm",color="red")+labs(x="Size class",y="Mean Language test score")+ theme(axis.title=element_text(size=11,face="bold"))
  

```


# Task 3


Now we want to add the relationship between score and social-economic status of pupil's family (SES). For doing that, we will cut this variable in order to show it as a categorical variable. For doing that, we use cut_interval() and we create three intervals of social-economic status: Low, medium and high. We will use the original data (without grouping) because we want to see the effects in the individual observations. 

In figure 3 we can see this relationship. Overall, we see that the biggest difference per type of class is in low economical status, and the smallest is in high economical status. Students with low economical status in a single-grade class will have better scores than the ones in multi-grade class, no matter the size of the class. Conversely, for high economical status, there is no significant difference between the scores for single-grade class and multi-grade class. 

Also, we see that in general, students of high economical status have higher scores than other socio economical statuses. But it is also important to notice that for high economical status the variance is much more pronounced for multi-grade than for single grade. 


```{r,fig.height=5,fig.width=7,fig.cap="Mean language test scores for size class, type of class and social-economic status of students in Neteherlands"}

nlschools4<-nlschools2 %>%
mutate("SES_level"=cut_interval(SES,3,labels=c("Low","Medium","High"))) 

ggplot(nlschools4,aes(class_size,lang),)+geom_point(position=position_jitter(),alpha=0.3)+facet_grid(vars(SES_level),vars(type_class))+geom_smooth(formula=y~x,method = "lm",color="red")+labs(x="Size class",y="Language test score")+ theme(axis.title=element_text(size=11,face="bold"))



```

```{r, echo=FALSE, include = FALSE}

ggplot(nlschools4,aes(class_size,lang,color=SES_level))+geom_point(position=position_jitter(),alpha=0.3)+facet_grid(vars(type_class))+geom_smooth(formula=y~x,method = "lm",color="red")+labs(x="Size class",y="Language test score")

```


To conclude, we see that the type of class has an impact on the test scores of the students. The best scores are obtained in single-grade for a small size class (between 10-20 students). Also, we see that socio-economical status have an impact in the language test scores of the students. High economical-status students tend to have higher scores, and also less differences between the type of class. The lowest scores are the students with low economical status and multi-grade class, and the scores with the highest dispersion are the students with high economical status and multi-grade class. Overall, the best scores are obtained by students with high-economical status in single-grade classes and small size classes. 


